---
title: "Fast and reliable use of numerical solvers"
date: "`r Sys.Date()`"
author: "Juho Timonen and Ben Bales"
link-citations: true
output:
  html_document:
    number_sections: true
    toc: true
    toc_float: true
    toc_depth: 4
    highlight: pygments
    theme: cosmo
    css: "style.css"
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load, message=FALSE}
# Requirements
require(rstan)
library(tidyverse)
require(bayesplot)
require(ggplot2)
require(loo)
require(stats)
require(posterior)
set.seed(123) # rng seed for data simulation
```

The required `posterior` package is not in CRAN but can be installed via 
`install.packages("posterior", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))`.

# Introduction

Assume we have a Bayesian model that involves a numeric solution to some ODE or
PDE (ordinary or partial differential equation), which contains
uncertain parameters. Try as we might, these calculations are often expensive
and frequently become the limiting factors in whatever model they are involved
in. The simplest things we might do to speed up our calculations are lowering
the timestep, coarsening the discretization, or increasing the tolerance of the
solvers.

That immediately leaves us with the question, is this okay? Has loosening the
tolerance affected the statistics of our solution? The technical answer is yes,
the practical answer is maybe, and the ideal case is no, but without checking
there's no way of knowing.

How can something like this be checked? The numerical approximation
answer is we should compare to a reference model and verify that the solution
at all points in space/time in the approximation are appropriately close
to the reference. That isn't so much of a problem in and of itself, but we're
doing statistics, and so we need to know that the solution is accurate enough
across all relevant parts of parameter space. Additionally, the "relevant parts
of parameter space" is the posterior itself which is not known beforehand!

The problem of validating an approximation in a Bayesian model is
significantly more complicated than in the classical numerical analysis world.
The point of this case study is to show how by adding one additional tool,
namely Pareto-Smoothed Importance Sampling (PSIS; @yao2018, @vehtari2019), we
can solve this problem.

# Workflow

Assume we have two models, $M_{low}$ and $M_{high}$.
These are otherwise the same model, except that $M_{low}$ uses a cheaper and
less accurate numerical methods (or just looser tolerances) to compute the
required ODE or PDE solutions. Therefore they have different posteriors,
$p_{low}$ and $p_{high}$, respectively.

To understand how PSIS comes into play, we must first discuss importance
sampling. If we want to compute expectations with the high precision model, we
can take draws from the low precision models and reweight these according to the
importance weights $\frac{p_{high}}{p_{low}}$. If these models are too
different, then the reweighting will produce noisy estimates that are not
useful. PSIS and particularly the $\hat{k}$ diagnostic (pronounced k-hat) is
the tool that tells us when we can or cannot rely on the importance weights. If
$\hat{k} < 0.5$ we are safe to do the importance sampling, if $\hat{k} < 0.7$
the importance sampling will start to converge more slowly, and if
$\hat{k} > 0.7$ the importance sampling estimates are unreliable. For simplicity
we will only consider the $\hat{k} < 0.5$ threshold.

Ideally, $M_{high}$ would involve a numerical method that we can trust
completely in all parts of the parameter space so that, as long as
$\hat{k} < 0.5$, importance weights can be used to reweight the low precision
approximation $p_{low}$ to the high precision approximation $p_{high}$. We can
think of $M_{high}$ as a reference model, because it is the baseline to which
we compare. It is difficult in practice to have a reference model that works
in all parts of parameter space because the properties of solvers (such as
stiffness of an ODE) can be very different in different parts of the parameter
space. Accuracy can be checked at a given set of parameters fairly easily,
but not over a high dimensional parameter space. Under these conditions it is
necessary to compromise to develop a reference model that works only over a
range of parameter space, but even then it is hard to know a-priori what range
that is.

We propose a workflow which addresses not only the problem of picking an
appropriate approximation but also the problem of developing a good reference
model.

::: {#workflow .box}
1. Generate draws from $p_{low}$.
2. Tune the numerical method in $M_{high}$ so that it is reliable at
these draws. All application specific knowledge and classical numerical
analysis can be used here.
3. Compute importance weights $\frac{p_{high}}{p_{low}}$
4. Compute the $\hat{k}$ diagnostic. If $\hat{k} > 0.5$, raise precision of $p_{low}$ and go back to step 1.
5. Resample draws using importance weights.
:::

The next two sections of this case study outline how to apply this workflow to:

1. Do inference with a low-precision ODE model and check and correct it
with a high precision solver after the fact
2. Do inference using a PDE solver that does not have explicit tolerance
controls

The importance sampling diagnostics are handled with the
[`loo`](https://mc-stan.org/loo) package and the resampling is handled with the
[`posterior`](https://github.com/stan-dev/posterior) package.

# SIR example (ODE)

Here we study a simple Susceptible-Infected-Recovered (SIR) model of disease
spread. This model is described in more detail in
[this case study](https://mc-stan.org/users/documentation/case-studies/boarding_school_case_study.html).

For the purposes of this case study, the goal is to use a very low
precision ODE solver to do inference and then check it afterwards against
a high precision solver. This can be done in practice if sampling with the
high precision solver itself takes an inordinate amount of time.

The states of the ODE are amounts of susceptible (S), infected (I) and recovered
(R) people. The dynamics are given by the ODE system:

\begin{align}
    \frac{dS}{dt} &= -\beta \cdot I \cdot \frac{S}{N} \\
    \frac{dI}{dt} &=  \beta  \cdot I \cdot \frac{S}{N} - \gamma \cdot I \\
    \frac{dR}{dt} &=  \gamma \cdot I
\end{align}

The parameters $\beta$ and $\gamma$ will be estimated from time series
observations of the number of infected people (I).

## Solving the system

First, we will import a function from `stan/sir.stan` for solving the ODE.
It uses the `integrate_ode_rk45` function that is built into Stan.

```{r sir_model, message=FALSE, results=FALSE}
model = stan_model("stan/sir.stan")
expose_stan_functions(model)
```

Second, we define various useful constants for our problem (the initial
conditions, etc.) and a wrapper function for the Stan function that solves
the ODE. The actual function exported from Stan is a bit awkward so we rewrap
it here in a way that the only options we will expose are parameters and
tolerance arguments.

```{r sir_solve}
N = 16                                # number of days measured
M = 1000                              # population size
I0 = 20                               # number of infected on day 0
initial_conditions = c(M - I0, I0, 0) # S, I, R on day 0
ts = seq_len(N)                       # measurement times
theta_true = c(1, 0.2)                # true parameter values

# Solve the SIR system and format result array
# - theta = [beta, gamma]
solve_sir = function(theta, rtol, atol, max_num_steps) {
  stan_solve_sir(initial_conditions, ts, theta,
                 c(0.0), M, rtol, atol, max_num_steps)
}
```

The ODE solvers in Stan are adaptive time step solvers, which estimate the
local error of the solution and adapt their step size so that the local error
estimates are less than `atol + rtol * abs(y)`, where `y` is the ODE solution,
and `atol` and `rtol` are called absolute and relative tolerance,
respectively. These tolerances need to be given, and affect both the accuracy
and computational cost of the solution. In general, `rtol` is the tolerance on
the relative error the solver can make when `y` is far from zero. When `abs(y)`
is small (of the order of `atol` or smaller), there is no need to achieve the
relative tolerance.

A third control parameter, `max_num_steps`, determines the maximum number of steps that can be taken to achieve the tolerance. 

*TODO:* add here what happens if max_num_steps is reached during sampling?
Reject? When the tolerances are set loosely, the sampler can become really
slow and hang totally if max_num_steps is large. My guess would be that this happends because it struggles to get out of a bad initial point as the
gradients are so wrong and it needs to do a lot of solves in places where it
will need to take a huge number of steps all the time. I don't
know but somehow this seems to be cured by setting max_num_steps lower and
pareto-k looks good. We
should do something like record the number of taken steps in different parts
of the parameter space.

We can just pick a couple tolerances and quickly run and plot a solution
just to get a feel for what the system looks like:

```{r sir_plot}
plot_sir = function(y) {
  y %>%
    as_tibble() %>%
    setNames(c("I")) %>%
    mutate(Day = ts) %>%
    ggplot(aes(Day, I)) +
    geom_line() +
    geom_point() +
    ylab("Infected people")
}

plot_sir(solve_sir(theta_true, 1e-4, 1e-4, 100))
```

In this model we are making noisy measurements of the number of infected
people (I) at each day.

## Generating test data

If we're going to generate data from our model we better have an accurate ODE
solver, otherwise we're just generating data from some weird approximate model.

The simplest way to check that an `atol` and `rtol` are suitable is to do
a solve at one tolerance level, repeat the solve at a much smaller (more
precise) tolerance, and then look at the maximum absolute error at any output
point. We will create a function to do this automatically:

```{r sir_check}
check_reliability = function(theta, rtol, atol, max_num_steps) {
  y_hat = solve_sir(theta, rtol, atol, max_num_steps)
  y_hat2 = solve_sir(theta, rtol / 10, atol / 10, max_num_steps)
  max_abs_err = max(abs(y_hat - y_hat2))
  return(max_abs_err)
}
```

We can study the maximum absolute error compared to a solution with 10 smaller
tolerances, as a function `tol = atol = rtol`.

```{r sir_tols, fig.width=7, fig.height=3.5}
errors = c()
tols = 10^(-c(1:12))
for (tol in tols) {
  errors = c(errors, check_reliability(theta_true, tol, tol, 1e8))
}

qplot(tols, errors, geom = c("point", "line")) +
  scale_x_log10() +
  scale_y_log10()
```

From this and our prior knowledge of infectious diseases, we assert
that $10^{-6}$ is a good enough value to use for `atol` and `rtol` during
simulation. Certainly we do not expect have a count of the infected
population accurate to $10^{-4}$ people.

We assume a negative binomial measurement model to get an observed number
of infected people (cases) at each time point.

```{r sir_noise, fig.width=7, fig.height=4.8}
atol =  1e-6
rtol = 1e-6

dispersion = 5 # noise parameter for negative binomial
mu = solve_sir(theta_true, atol, rtol, 1e8)
y = stats::rnbinom(length(ts), mu = mu, size = dispersion)

tibble(t = ts, mu = mu, y = y) %>%
  ggplot() +
  geom_line(aes(t, mu), col = "firebrick") +
  geom_point(aes(t, y)) +
  xlab("Day") +
  ylab("Infected people") +
  ggtitle("Simulated data as points \nUnderlying solution as lines")
```

## Applying the workflow

As a reminder, our mission in fitting this ODE is to use a low precision
solver. It is always tempting to use low precision solvers when working
with ODEs because they run faster. The difficulty becomes how to deal with
the approximation. Does the approximation matter? If so, how can it be
corrected? These are the questions the workflow here will allow us to answer.

### Generating draws from $p_{low}$

The first step in the workflow is to take any low precision approximation and
fit the data. Remember, all our numerical methods are approximations, and so
we refer to this model specifically as a low precision model. We will check it
against a higher precision model later. In this case, we will use
`rtol = 1e-4`, `atol = 1e-3`.

```{r sir_fit, fig.width=7, fig.height=4.5, results=FALSE}
rtol_low = 1e-4
atol_low = 1e-3
max_num_steps_low = 100

fit = rstan::sampling(model, list(
  N = length(ts),
  M = M,
  ts = ts,
  y = y,
  initial_conditions = initial_conditions,
  rtol = rtol_low,
  atol = atol_low,
  max_num_steps = max_num_steps_low))
```


```{r sir_summary}
print(fit, pars = c("beta", "gamma", "phi"))
```

### Developing a reference model

Before we can check if the importance sampling correction is possible, we
need to have a reference model to compare against. That means we need a
version of the model with tolerances such that it is suitably accurate across
all the posterior draws generated from the low precision model.

In this case, `rtol = atol = 1e-6` was accurate enough generating the data,
so let's check if it is accurate in all the draws in this posterior.

```{r sir_tune}
atol_high = 1e-6
rtol_high = 1e-6
max_num_steps_high = 1e6
draws = rstan::extract(fit, pars = c("beta", "gamma", "phi"))
phi_draws = draws$phi
theta_draws = cbind(draws$beta, draws$gamma)
num_draws = length(phi_draws)

# Compute differences
errors = c()
for (i in 1:num_draws) {
  mae = check_reliability(theta_draws[i,], atol_high, rtol_high,
                          max_num_steps_high)
  errors = c(errors, mae)
}
```

We can plot this as a distribution and see that `rtol = atol = 1e-6` keeps
us under an absolute error of one milliperson. This seems accurate enough.

```{r sir_hist}
qplot(errors, geom = "histogram")
```

### Computing importance weights

With the reference model in place, it is time to compute the importance
weights $\frac{p_{high}}{p_{low}}$. This is simple:
just compute the log density of the reference model and the log
density of the low precision model and take the difference (we work
with the log of the importance ratios $\log p_{high} - \log p_{low}$
for numeric stability).

The hidden downside is that it might take some time to compute the log
densities of the reference model. The upside is that all the calculations
can be done in parallel, and it should still be way faster than sampling
with the reference model itself.

If the priors are kept the same between the reference and low precision
model, then those can be left out of this calculation (they will cancel).

```{r sir_weights}
log_lh_low = rep(0, num_draws)
log_lh_high = rep(0, num_draws)
for (i in seq_len(num_draws)) {
  y_hat_low = solve_sir(theta_draws[i,], rtol_low, atol_low,
                        max_num_steps_low)
  y_hat_high = solve_sir(theta_draws[i,], rtol_high, atol_high, 
                         max_num_steps_high)
  log_lh_low[i] = sum(dnbinom(y, size = phi_draws[i],
                              mu = y_hat_low, log = TRUE))
  log_lh_high[i] = sum(dnbinom(y, size = phi_draws[i],
                              mu = y_hat_high, log = TRUE))
}
log_weights = log_lh_high - log_lh_low
```

We can plot the log importance ratios and see they are all close to zero
(which means out approximation was not too bad).

```{r sir_weight_hist}
qplot(log_weights, geom = "histogram")
```

### Computing $\hat{k}$ diagnostic

With the importance ratios calculated, we can check if they are usable or not
with the PSIS $\hat{k}$ diagnostic.

```{r sir_pareto}
loo::psis(log_weights)
```

$\hat{k} < 0.5$, and so importance sampling should be reliable.

### Resampling

At this point we have a weighted set of posterior draws. It
is usually easier to work with a set of draws than a set of weighted draws, so
we can resample our weighted draws to become a set of unweighted draws using
`posterior::resample_draws`. The effective sample size will be slightly
lowered by such a resampling, but unweighted draws are really more convenient
to work with.

Just because it is possible to do an importance sampling correction on a set
of draws does not mean that unweighted statistics on these draws are safe
to use. In this case, the results are not much different, but it should not
be forgotten:

```{r sir_resample}
w = exp(log_weights)
draws_list = rstan::extract(fit, c("beta", "gamma", "phi"))
draws_df = posterior::as_draws_df(draws_list)
resampled_df = posterior::resample_draws(draws_df,
                                         weights = w)

print(draws_df %>% posterior::summarize_draws())
print(resampled_df %>% posterior::summarize_draws())
```

<!-- WEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEE -->

# Heat Equation Example (PDE)

In this example we consider the diffusion of heat ($u(t, x)$) in a rod
($x \in [0, L]$).

For the purposes of this case study, the goal is to use a PDE solver in a
model that has no automatic error control, only fixed discretization controls.
This problem comes up from time to time when problems demand a custom solver
be written -- it is not always easy to tack on error control algorithms.

In this hypothetical experiment, the rod is cooled to room
temperature and then heated from the left side. After some time the temperature
profile of the rod is measured and from this the thermal diffusivity will be
estimated.

The dynamics are governed by the 1D heat equation:

\begin{align}
\frac{\partial u}{\partial t} &= K \cdot \frac{\partial^2 u}{\partial x^2} \\
u(0, x) &= 0 \\
u(t, 0) &= 1 \\
u(t, L) &= 0
\end{align}

## Solving the system
All of the computations in this example are going to be done with a method of
lines discretization of this problem and a backwards Euler integrator. The
appropriate math is described in the online lecture notes
[ATM 623: Climate Modeling](http://www.atmos.albany.edu/facstaff/brose/classes/ATM623_Spring2015/Notes/Lectures/Lecture16%20--%20Numerical%20methods%20for%20diffusion%20models.html) by Brian E. J. Rose, though any introductory PDE reference should suffice.

For convenience we have defined a Stan function that solves equations above
and computes the measured temperatures in the system given a timestep, a
spatial discretization, a hypothetical diffusivity, a measurement time, and a
list of measurement points.

```{r diffusion_model, message=FALSE, results=FALSE}
model = stan_model("stan/diffusion.stan")
expose_stan_functions(model)
```

```{r diffusion_solve}
dt = 1.0
Nx = 10
K = 1e-1
T_meas = 0.1
x_meas = c(-1.0, 0.01, 0.5, 0.99, 1.0, 2.0)

solve_pde(dt, Nx, K, T_meas, x_meas)
```

The function has the signature:

```
vector solve_pde(dt, Nx, K, T_meas, x_meas)
```

with arguments:

* `dt` - Timestep
* `Nx` - Number of interior points in spatial discretization
* `K` - Thermal diffusivity
* `T_meas` - Measurement time
* `x_meas` - Measurement points

Assume a true thermal diffusivity $K_{true} = 0.05$ and that we measure the
temperature in the rod at `Nx` points evenly spaced on the rod. We will
generate data under these conditions and try to recover the diffusivity later.

First, let's set up constants and plot a possible solution with measurement
points:

```{r diffusion_setup, fig.width = 6.75, fig.height = 5}
dt = 1e-1
Nx = 5
L = 1.0

x_meas = seq(0.0, L, length = 7)[2:6]
T_meas = 1.0
K_true = 0.015

# For these calculations pretend we are measuring everywhere so we can
#  see the whole solution
x = seq(-0.1, 1.1, length = 100)
u0 = c(rep(1.0, sum(x <= 0.0)), rep(0.0, sum(x > 0.0)))
uT = solve_pde(dt, Nx, K_true, T_meas, x)

# Solve at only the measurement points
mu = solve_pde(dt, Nx, K_true, T_meas, x_meas)

# Plot
tibble(x = x, `u(t = 0)` = u0, `u(t = T)` = uT) %>%
  gather(Legend, u, -x) %>%
  ggplot(aes(x, u)) +
  geom_line(aes(color = Legend, group = Legend)) +
  geom_point(data = tibble(x = x_meas, u = mu)) +
  ggtitle("Measurement points indicated in black\nDashed lines indicate boundary of sample") +
  geom_vline(aes(xintercept = 0.0), linetype = "dashed") +
  geom_vline(aes(xintercept = L), linetype = "dashed")
```

The red line shows the initial conditions. Because the solution is actually
discretized, to only five points on the rod (seven including the boundaries)
we do linear interpolation to get the values in the intermediate points (which
makes the boundary look a bit strange).

The teal points show the distribution of heat in the rod at time `t = T`, where
we plan to take measurements (indicated by the black dots) and make an
inference about the unknown thermal diffusivity of the rod.

## Generating test data

Now that we can compute solutions to this problem, our first question will
be is a given solution accurate enough? The simple way to check this is by
computing the solution again at a higher space/time resolution and checking
the difference.

We can define a convenience function that for a given discretization and
experimental configuration computes a solution and also another solution at
higher precision and returns the maximum absolute error.

```{r diffusion_check}
# Function to help determine if dt and Nx are small enough at given K
check_reliability = function(dt, Nx, K, T_meas, x_meas) {
  mu = solve_pde(dt, Nx, K, T_meas, x_meas)
  mu_more_accurate = solve_pde(dt / 2.0, 2 * Nx, K, T_meas, x_meas)
  max_abs_err = max(abs(mu_more_accurate - mu))
  return(max_abs_err)
}

# Check at K = K_true
check_reliability(dt, Nx, K_true, T_meas, x_meas)
```

Is that error good or is that bad? That is something that will need to be
determined in the context of the application. In this case we are going assume
a measurment noise of $0.1$, and so we should get our numerical error quite a
bit below that.

```{r diffusion_check2}
dt = 0.01
Nx = 40
check_reliability(dt, Nx, K_true, T_meas, x_meas)
```

This seems good enough for now, but you might further refine your solution. Now
to simulate data:

```{r diffusion_noise}
sigma = 0.1
noise = rnorm(length(x_meas), 0, sigma)
y = solve_pde(dt, Nx, K_true, T_meas, x_meas) + noise
```

## Applying the workflow

### Generating draws from $p_{low}$

Now that we have simulated data, it is time to do inference. The first step,
similarly as for the ODE, is to fit an initial approximate model to our
data. Again, all our calculations are approximations, and so we refer to this
model as a low precision model because we will check it against a higher
precision model later.

Assume we are very impatient and want this computation to finish quickly so
that we use only one timestep and one spatial point of resolution in our
discretization:

```{r diffusion_fit1, results=FALSE}
dt_low = 1.0
Nx_low = 1
fit = rstan::sampling(model,
                data = list(dt = dt_low,
                            Nx = Nx_low,
                            N_meas = length(x_meas),
                            T_meas = T_meas,
                            x_meas = x_meas,
                            y = y),
                control = list(adapt_delta = 0.95),
                cores = 1)
```

Let us look at our results:

```{r diffusion_summary1}
print(fit, pars = c("K", "sigma"))
```

We remember from earlier that $K_{true} = 0.1$, $\sigma_{true} = 0.1$,
so something is off. We will diagnose this using our approximation tools.

### Developing a reference model

Again, to check if the importance sampling correction can be done, we need
a reference model that works for all the posterior draws we got from the
low precision model. We can develop the reference model using the same
technique we did previously (guess a high precision, and check the maximum
absolute error between that high precision model and one of even higher
precision).

```{r diffusion_tune1}
dt_high = 0.01
Nx_high = 100

K_draws = rstan::extract(fit, "K")$K
sigma_draws = rstan::extract(fit, "sigma")$sigma
num_draws = length(K_draws)

# Compute differences
errors = c()
for (i in 1:num_draws) {
  mae = check_reliability(dt_high, Nx_high, K_draws[i], T_meas, x_meas)
  errors = c(errors, mae)
}
```

With a simple one parameter model we can plot our approximate errors as a
function of K (so we know the solution is suitable everywhere).

```{r diffusion_errors1}
error_plot = function(K, mae) {
  ylab = 
  df = data.frame(K, mae)
  ggplot(df, aes_string(x = "K", y = "mae")) + 
    geom_point(col = "#1864de", alpha = 0.5) +
    xlab("K") + ylab("Max. absolute error")
}
error_plot(K_draws, errors)
```

The errors here seem low enough.

### Computing importance weights

The importance weights $\frac{p_{high}}{p_{low}}$  are computed on the log
scale. The priors cancel out so we only need to work with log likelihoods.

Again, this step looks simple in this example, but in practice it might be
more complicated. It is possible that the reference calculation is done
with an entirely different piece of software. For instance, with a PDE
perhaps a the reference solution is computed with a well-tested FEM solver
in a different software environment entirely.

```{r diffusion_weights1}
# Define a function
compute_log_weights = function(
  dt_low, Nx_low, dt_high, Nx_high, 
  K_draws, T_meas, x_meas, y_meas)
{
  log_lh_low = rep(0, num_draws)
  log_lh_high = rep(0, num_draws)
  for (i in seq_len(num_draws)) {
    mu_low = solve_pde(dt_low, Nx_low, K_draws[i], T_meas, x_meas)
    mu_high = solve_pde(dt_high, Nx_high, K_draws[i], T_meas, x_meas)
    log_lh_low[i] = sum(dnorm(y_meas, mu_low, sigma_draws[i], log = TRUE))
    log_lh_high[i] = sum(dnorm(y_meas, mu_high, sigma_draws[i], log = TRUE))
  }
  log_weights = log_lh_high - log_lh_low
  return(log_weights)
}

# Apply function
log_weights = compute_log_weights(dt_low, Nx_low, dt_high, Nx_high,
                                  K_draws, T_meas, x_meas, y)
```

### Computing $\hat{k}$ diagnostic

If the $\hat{k}$ diagnostic is not low enough, it is not possible to do the
importance sampling correction and we need to recompute our posterior with a
higher resolution model. The `loo` package computes the $\hat{k}$ diagnostic
for us:

```{r diffusion_pareto1, fig.width=6, fig.height=4.5}
loo::psis(log_weights)
```

Oh no! $\hat{k} > 0.5$, and it turns out modeling this process with one timestep
and one spatial point was not a good idea. This means we need to up the precision
in the low resolution model and go back to *Step 1*.

### Repeating the loop

```{r diffusion_fit2, results=FALSE}
dt_low = 0.1
Nx_low = 10
fit = rstan::sampling(model,
                data = list(dt = dt_low,
                            Nx = Nx_low,
                            N_meas = length(x_meas),
                            T_meas = T_meas,
                            x_meas = x_meas,
                            y = y),
                control = list(adapt_delta = 0.95),
                cores = 1)
```

Again, we can check our regular diagnostics:

```{r diffusion_summary2}
print(fit, pars = c("K", "sigma"))
```

Again, we verify our reference solution:

```{r diffusion_tune2}
K_draws = rstan::extract(fit, "K")$K
sigma_draws = rstan::extract(fit, "sigma")$sigma
num_draws = length(K_draws)

# Compute differences
errors = c()
for (i in 1:num_draws) {
  mae = check_reliability(dt_high, Nx_high, K_draws[i], T_meas, x_meas)
  errors = c(errors, mae)
}

# Plot
error_plot(K_draws, errors)
```

And again we can compute the importance ratios and run the PSIS diagnostics on
them:

```{r diffusion_weights2}
log_weights = compute_log_weights(dt_low, Nx_low, dt_high, Nx_high,
                                  K_draws, T_meas, x_meas, y)
loo::psis(log_weights)
```

And this time $\hat{k} < 0.5$, so we are good enough!

### Resampling draws

At this point we have a weighted set of posterior draws. Again it
is usually easier to work with a set of draws than a set of weighted draws, so
we resample our weighted draws using `posterior::resample_draws`.

```{r diffusion_resampling}
w = exp(log_weights)
draws_list = rstan::extract(fit, c("K", "sigma"))
draws_df = posterior::as_draws_df(draws_list)
resampled_df = posterior::resample_draws(draws_df,
                                         weights = w)

print(draws_df %>% posterior::summarize_draws())
print(resampled_df %>% posterior::summarize_draws())
```

<!-- WEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEEE -->

# Conclusions

And that is that! Happy approximating!

# Computation environment

```{r computational_environment}
sessionInfo()
```

# References
