---
title: "Fast and reliable use of numerical solvers"
date: "`r Sys.Date()`"
author: "Juho Timonen and Ben Bales"
link-citations: true
output:
  html_document:
    number_sections: true
    toc: true
    toc_float: true
    toc_depth: 4
    highlight: pygments
    theme: cosmo
    css: "style.css"
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load, message=FALSE}
# Requirements
require(rstan)
library(tidyverse)
require(bayesplot)
require(grDevices)
require(ggplot2)
require(loo)
require(stats)
require(posterior)
set.seed(123)
```

The required `posterior` package is not in CRAN but can be installed via 
`install.packages("posterior", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))`.

# Introduction

Assume we have a Bayesian model that involves a numeric solution to some ODE or PDE (ordinary or partial differential equation), which contains
uncertain parameters. Try as we might, these calculations are often expensive and frequently become the limiting factors in whatever model they are involved in. The simplest things we might do to speed up our calculations are lowering the timestep, coarsening the discretization, or increasing the tolerance of the solvers.

That immediately leaves us with the question, is this okay? Has loosening the
tolerance affected the statistics of our solution? The technical answer is yes,
the practical answer is maybe, and the ideal case is no, but without checking
there's no way of knowing.

And so how can something like this be checked? The numerical approximation
answer is we should compare to a reference model and verify that the solution
at all points in space/time in the approximation are appropriately close
to the reference. That isn't so much of a problem in and of itself, but we're
doing statistics, and so we need to know that the solution is accurate enough
across all relevant parts of parameter space. Additionally, the "relevant parts
of parameter space" is the posterior itself which is not known beforehand!

And so the problem of validating an approximation in a Bayesian model is
significantly more complicated than in the classical numerical analysis world.
The point of this case study is to show how by adding one additional tool, namely Pareto-Smoothed Importance Sampling (PSIS; @yao2018, @vehtari2019), we can solve this problem.

# Workflow

Assume we have two models, $M_{low}$ and $M_{high}$.
These are otherwise the same model, except that $M_{low}$ uses a cheaper and less accurate numerical methods (or just looser tolerances) to compute the required ODE or PDE solutions. Therefore they have different posteriors, $p_{low}$ and $p_{high}$, respectively.

To understand how PSIS comes into play, we must first discuss importance sampling. If we want to compute expectations with the high precision model, we can take draws from the low precision models and reweight these according to the
importance weights $\frac{p_{high}}{p_{low}}$. If these models are too
different, then the reweighting will produce noisy estimates that are not
useful. PSIS and particularly the $\hat{k}$ diagnostic (pronounced k-hat) is
the tool that tells us when we can or cannot rely on the importance weights. If
$\hat{k} < 0.5$ we are safe to do the importance sampling, if $\hat{k} < 0.7$
the importance sampling will start to converge more slowly, and if
$\hat{k} > 0.7$ the importance sampling estimates are unreliable. For simplicity
we will only consider the $\hat{k} < 0.5$ threshold.

Ideally, $M_{high}$ would involve a numerical method that we can trust
completely, in all parts of the parameter space. This is required because for example the properties (such as stiffness) of an ODE system can be very different in different parts of the parameter space. The problem however, is that it is not generally possible to check whether a given method is reliable everywhere. Classical numerical methods can be used to check the reliability, but only for one parameter combination at a time.

We propose a workflow, which neatly solves not only the problem of picking an appropriate approximation, but also the problem of developing a good reference model.

::: {#workflow .box}
1. Generate draws from $p_{low}$.
2. Tune the numerical method in $M_{high}$ so that it is reliable at
these draws. All application specific knowledge and classical numerical
analysis can be used here.
3. Compute importance weights $\frac{p_{high}}{p_{low}}$
4. Compute the $\hat{k}$ diagnostic. If $\hat{k} > 0.5$, raise precision of $p_{low}$ and go back to step 1.
5. Resample draws using importance weights.
:::

The next section of this case study demonstrates how to apply this workflow on a simple PDE model using Stan in combination with the [`loo`](https://mc-stan.org/loo) and
[`posterior`](https://github.com/stan-dev/posterior) software packages.

# Heat Equation Example

## Problem definition
In this example we consider the diffusion of heat ($u(t, x)$) in a rod
($x \in [0, L]$). In this hypothetical experiment, the rod is cooled to room
temperature and then heated from the left side. After some time the temperature
profile of the rod is measured and from this the thermal diffusivity is to be
estimated.

The dynamics are governed by the 1D heat equation:

\begin{align}
\frac{\partial u}{\partial t} &= K \cdot \frac{\partial^2 u}{\partial x^2} \\
u(0, x) &= 0 \\
u(t, 0) &= 1 \\
u(t, L) &= 0
\end{align}


## Solving the system
All of the computations in this example are going to be done with a method of
lines discretization of this problem and a backwards Euler integrator. The
appropriate math is described in the online lecture notes
[ATM 623: Climate Modeling](http://www.atmos.albany.edu/facstaff/brose/classes/ATM623_Spring2015/Notes/Lectures/Lecture16%20--%20Numerical%20methods%20for%20diffusion%20models.html) by Brian E. J. Rose, though any introductory PDE reference should suffice.

For convenience we have defined a Stan function that solve our problem
given a timestep, a spatial discretization, a hypothetical diffusivity, a
measurement time, and a list of measurement points and returns the predicted
temperature at each of those measurement points.

```{r}
model = stan_model("diffusion/diffusion.stan")
expose_stan_functions(model)
```

```{r}
dt = 1.0
Nx = 10
K = 1e-1
T_meas = 0.1
x_meas = c(-1.0, 0.01, 0.5, 0.99, 1.0, 2.0)

solve_pde(dt, Nx, K, T_meas, x_meas)
```

The function has the signature:

```
vector solve_pde(dt, Nx, K, T_meas, x_meas)
```

with arguments:

* `dt` - Timestep
* `Nx` - Number of interior points in spatial discretization
* `K` - Thermal diffusivity
* `T_meas` - Measurement time
* `x_meas` - Measurement points

Assume a true thermal diffusivity $K_{true} = 0.05$ and that we measure the
temperature in the rod at `Nx` points evenly spaced on the rod. We will
generate data under these conditions and try to recover the diffusivity later.

First, let's set up constants and plot a possible solution:

```{r diffusion, fig.width = 6.75, fig.height = 5}
dt = 1e-1
Nx = 5
L = 1.0

# Pretend we're measuring everywhere so we see the full solution
x = seq(-0.1, 1.1, length = 100)
x_meas = seq(0.0, L, length = 7)[2:6]
T_meas = 1.0
K_true = 0.025

u0 = solve_pde(dt, Nx, K_true, 0.0, x)
uT = solve_pde(dt, Nx, K_true, T_meas, x)
mu = solve_pde(dt, Nx, K_true, T_meas, x_meas)

plot(x, u0, type = 'l')
lines(x, uT, col = 'firebrick')
points(x_meas, mu, col = '#1864de')
```

The blue points in the solution are the ones we compare to (these
are the parts of the solution we care most about). *TODO: this figure probably needs more explaining.*

## Reliability of the solution

Our first question will be: is this solution accurate enough at the blue points? The simple way to check this is by computing the solution at the blue points at a higher space/time resolution and checking the difference.

We define a function that we can use for determining reliability, given a
fixed parameter value $K$. In our approach we double the space and time resolution and check the maximum absolute error.

```{r}
# Function to help determine if dt and Nx are small enough at given K
check_reliability = function(dt, Nx, K, T_meas, x_meas) {
  mu = solve_pde(dt, Nx, K, T_meas, x_meas)
  mu_more_accurate = solve_pde(dt / 2.0, 2 * Nx, K, T_meas, x_meas)
  max_abs_err = max(abs(mu_more_accurate - mu))
  return(max_abs_err)
}

# Check at K = K_true
check_reliability(dt, Nx, K_true, T_meas, x_meas)
```

Is that good or is that bad? That is something that will need to be determined in the context of the application. In this case I am going to generate data with a standard deviation of $0.1$, and so I want to get my numerical error quite a
bit below that before I am comfortable.

```{r}
dt = 0.01
Nx = 40
check_reliability(dt, Nx, K_true, T_meas, x_meas)
```

This seems good enough for now, but you might further refine your solution. Now
to simulate data:

```{r}
sigma = 0.1
noise = rnorm(length(x_meas), 0, sigma)
y = solve_pde(dt, Nx, K_true, T_meas, x_meas) + noise
```

## Applying the workflow

### Generating draws from $p_{low}$

We now take off our data-generating hats and put on our model-fitting hats. We fit an initial approximate model to our data. All our calculations are
approximations, and so we refer to this model as a low resolution model, because
we will check the calculations we did with it in the following steps.

Pretend `x_meas` is measured data and we do not know the true thermal diffusivity. Perhaps we are greedy and want this computation to finish quickly and so we fit this with a very aggressive approximation (one timestep, one spatial point):

```{r}
dt_low = 1.0
Nx_low = 1
fit = rstan::sampling(model,
                data = list(dt = dt_low,
                            Nx = Nx_low,
                            N_meas = length(x_meas),
                            T_meas = T_meas,
                            x_meas = x_meas,
                            y = y),
                cores = 4)
```

Looking at the posterior it appears everything was successful and the chains
mixed. However, we remember from earlier that $K_{true} = 0.1$, so something is
off. We will diagnose this using our approximation tools though.

```{r}
print(fit, pars = c("K", "sigma"))
```

### Tuning the reference method

We tune the high resolution method the same way as we did when generating the data, except that now we need to check the maximum absolute error (MAE) for all
parameter values that we sampled.

```{r}
dt_high = 0.005
Nx_high = 80

K_draws = rstan::extract(fit, "K")$K
sigma_draws = rstan::extract(fit, "sigma")$sigma
num_draws = length(K_draws)

# Compute differences
errors = c()
for (i in 1:num_draws) {
  mae = check_reliability(dt_high, Nx_high, K_draws[i], T_meas, x_meas)
  errors = c(errors, mae)
}
```

With a simple one parameter model we can plot our approximate errors as a
function of K (so we know the solution is suitable everywhere).

```{r}
error_plot = function(K, mae) {
  ylab = "Max. absolute error"
  plot(K, mae, pch = 4, col = "#1864de", xlab = "K", ylab = ylab)
}
error_plot(K_draws, errors)
```

The errors here seem low enough.

### Computing importance weights

The importance weights $\frac{p_{high}}{p_{low}}$  are computed on the log
scale. The priors cancel out so we only need to work with log likelihoods.

```{r}
# Define a function
compute_log_weights = function(
  dt_low, Nx_low, dt_high, Nx_high, 
  K_draws, T_meas, x_meas, y_meas)
{
  log_lh_low = rep(0, num_draws)
  log_lh_high = rep(0, num_draws)
  for (i in seq_len(num_draws)) {
    mu_low = solve_pde(dt_low, Nx_low, K_draws[i], T_meas, x_meas)
    mu_high = solve_pde(dt_high, Nx_high, K_draws[i], T_meas, x_meas)
    log_lh_low[i] = sum(dnorm(y_meas, mu_low, sigma_draws[i], log = TRUE))
    log_lh_high[i] = sum(dnorm(y_meas, mu_high, sigma_draws[i], log = TRUE))
  }
  log_weights = log_lh_high - log_lh_low
  return(log_weights)
}

# Apply function
log_weights = compute_log_weights(dt_low, Nx_low, dt_high, Nx_high,
                                  K_draws, T_meas, x_meas, y)
```

### Computing $\hat{k}$ diagnostic

If the $\hat{k}$ diagnostic is not low enough, it is not possible to do the
importance sampling correction and we need to recompute our posterior with a
higher resolution model. The `loo` package computes the $\hat{k}$ diagnostic for us:

```{r pareto, fig.width=6, fig.height=4.5}
loo::psis(log_weights)
```

Oh no! $\hat{k} > 0.5$, and it turns out modeling this process with one timestep
and one spatial point is not a good idea. This means we need to up the precision in the low resolution model and go back to *Step 1*.

### Repeating the loop

```{r}
dt_low = 0.1
Nx_low = 10
fit = rstan::sampling(model,
                data = list(dt = dt_low,
                            Nx = Nx_low,
                            N_meas = length(x_meas),
                            T_meas = T_meas,
                            x_meas = x_meas,
                            y = y),
                cores = 4)
```

Again, we can check our regular diagnostics:

```{r}
print(fit, pars = c("K", "sigma"))
```

Again, we verify our reference solution:

```{r}
dt_high = 0.005
Nx_high = 80
K_draws = rstan::extract(fit, "K")$K
sigma_draws = rstan::extract(fit, "sigma")$sigma
num_draws = length(K_draws)

# Compute differences
errors = c()
for (i in 1:num_draws) {
  mae = check_reliability(dt_high, Nx_high, K_draws[i], T_meas, x_meas)
  errors = c(errors, mae)
}

# Plot
error_plot(K_draws, errors)
```

And again we can compute the importance ratios and run the PSIS diagnostics on
them:

```{r}
log_weights = compute_log_weights(dt_low, Nx_low, dt_high, Nx_high,
                                  K_draws, T_meas, x_meas, y)
loo::psis(log_weights)
```

And this time $\hat{k} < 0.5$, so we are good enough!

### Resampling draws

This is optional. At this point we have a weighted set of posterior draws. It
is usually easier to work with a set of draws than a set of weighted draws, so
we can resample our weighted draws using `posterior::resample_draws`.

The last step is resampling. We have draws of $K$ and $\sigma$, but they are
weighted, and usually it is easier to just use a set of draws. Just because it
is possible to do an importance sampling correction on a set of draws does not
mean that unweighted statistics on these draws are safe to use. In this case,
the results are not much different, but it is relatively trivial to do a
resampling with the `posterior` package:

```{r}
w = exp(log_weights)
draws_list = rstan::extract(fit, c("K", "sigma"))
draws_df = posterior::as_draws_df(draws_list)
resampled_df = posterior::resample_draws(draws_df,
                                         weights = w)

print(draws_df %>% posterior::summarize_draws())
print(resampled_df %>% posterior::summarize_draws())
```

# Computation environment

```{r session}
sessionInfo()
```

# References
