---
title: "Reliable and efficient numerical integration of ODEs"
date: "`r Sys.Date()`"
author: "Juho Timonen, Ben Bales, Harri Lähdesmäki and Aki Vehtari"
link-citations: true
output:
  html_document:
    number_sections: true
    toc: true
    toc_float: true
    toc_depth: 4
    highlight: pygments
    theme: cosmo
    css: "style.css"
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load, message=FALSE}
# Requirements
require(cmdstanr)
require(bayesplot)
require(ggplot2)
require(loo)
require(stats)

rng_seed_r <- 999 # rng seed for MCMC
rng_seed_stan <- 123 # rng seed for data simulation
set.seed(rng_seed_r)
```

# Introduction

Ordinary differential equations (ODEs) 

\begin{equation}
\frac{\text{d} \textbf{y}(t)}{\text{d} t} = f_{\theta}(\textbf{y}(t), t)
\end{equation}

are a model for the rate of change in in some quantities $\textbf{y}(t)$,
and are used in many fields of science. It is common that we have
time-course data $\textbf{y}^{\dagger}_i$, observed 
at time points $t_i \neq t_0$, and we need to calculate $\textbf{y}(t_i)$ in
order to compute the likelihood. Given initial value 
$\textbf{y}(t_0) = \textbf{y}_0$, and assuming some smoothness requirements for
$f_{\theta}$, there exists a unique solution for
$\textbf{y}(t)$, $t \in \mathbb{R}$ but it often has no closed form. Numerical methods are then required.

When performing Bayesian inference for the parameters $\theta$
(and possible other model parameters) using Stan,
the system needs to be solved numerically on each log posterior probability
evaluation. Furthermore, Stan needs to compute gradients for these solutions.
Try as we might, these computations are often expensive and frequently become
the limiting factors in whatever model they are involved in. The same problem
can occur also with any other type of model that requires numerically
solving an implicitly defined function or variable, but in this case study
we focus on ODEs.

The numerical methods for ODEs usually involve some discretization in
time, which affects the solver accuracy. Denser discretization means
more accuracy but also more computation. Alternatively, methods can estimate
their error and adapt their step size or other properties so that a given tolerance is achieved. The latter is what the built-in ODE solvers in Stan do.
The relationship between tolerance and runtime is 

The simplest things we might do to speed up our calculations are lowering
the timestep, coarsening the discretization, or increasing the tolerance of the
solvers. That immediately leaves us with the question, is this okay? Has
changing the numerical method affected our parameter inference results? Was our
original method giving correct inference results to begin with? Are the default
tolerances in Stan suitable for the problem at hand?

The solution provided by a numerical method is always an approximation to the
true solution $\textbf{y}(t)$. This is why our
posterior probability density evaluations are also approximations and
the whole MCMC inference can be thought to be biased to some degree. However,
we can think that the inference results are correct if making the numerical
method more accurate does not affect the statistical properties of the
posterior draws.

How can something like this be checked? The first problem is that it might not
be computationally very attractive to run sampling repeatedly, gradually
increasing the solver accuracy. If the model parameters are fixed,
we can verify that the solution at all points in space/time is
appropriately close to a more accurate reference solution. That isn't so much
of a problem in and of itself, but we are doing statistics, and so we need to
know that the solution is accurate enough across all relevant parts of
parameter space. Additionally, it is not known beforehand where the 
"relevant parts of parameter space" are!

The problem of validating the use of a numerical method for a Bayesian model
is therefore significantly more complicated than in the classical numerical
analysis world. The point of this case study is to show how by adding one
additional tool, namely Pareto-Smoothed Importance Sampling (PSIS) [@yao2018;
@vehtari2019], we can solve this problem.

# Workflow

Let $M$ be the model for which we would like to perform inference, but which we
cannot evaluate since the likelihood is defined implicitly through an
ODE or PDE system that is not analytically tractable. MCMC inference for $M$
can be seen actually as inference for another model $M_{high}$, which
is the same model as $M$ but using a numerical solver, and can therefore be
evaluated.

Our workflow addresses the problem of defining the high-precision numerical
method in $M_{high}$ so that $M_{high}$ can trusted to have essentially the
same posterior as $M$. We define a way to perform inference for $M_{high}$ 
without needing to compute gradients or HMC trajectories for it. This involves
another model $M_{low}$, which is again the same model, except that $M_{low}$
uses a cheaper and less accurate numerical methods (or just looser tolerances
and/or coarser discretization) to compute the required ODE or PDE solutions,
and is therefore faster to fit. The posterior densities are denoted $p_{low}$
and $p_{high}$, respectively.

To understand how PSIS comes into play, we must first discuss importance
sampling. If we want to compute expectations with the high precision model, we
can take draws from the low precision models and reweight these according to the
importance weights $\frac{p_{high}}{p_{low}}$. If these models are too
different, then the reweighting will produce noisy estimates that are not
useful. PSIS and particularly the Pareto $k$-diagnostic (denoted $\hat{k}$), is
the tool that tells us when we can or cannot rely on the importance weights. If
$\hat{k} < 0.5$ we are safe to do the importance sampling, if $\hat{k} < 0.7$
the importance sampling will start to converge more slowly, and if
$\hat{k} > 0.7$ the importance sampling estimates are unreliable. For simplicity
we will only consider the $\hat{k} < 0.5$ threshold.

Ideally, $M_{high}$ would involve a numerical method that we can trust
completely in all parts of the parameter space so that, as long as
$\hat{k} < 0.5$, importance weights can be used to reweight the low precision
approximation $p_{low}$ to the high precision approximation $p_{high}$. We can
think of $M_{high}$ as a reference model, because it is the baseline to which
we compare. It is difficult in practice to know if a given model is a good
reference model in all parts of parameter space, due to curse of dimensionality
and the fact that analysed system can have different properties in different
parts of the parameter space. For example, ODEs can qualitatively change their
behaviour as a function of parameters (bifurcation), or become stiff or
chaotic in some parameter regions. Accuracy can be checked at a given set of
parameters fairly easily, but not over a high dimensional parameter space.
Under these conditions it is necessary to compromise to develop a reference
model that works only over a range of parameter space, but even then it is hard
to know *a priori* what range that is.

We propose the following workflow:

::: {#workflow .box}
1. Generate draws from $p_{low}$.
2. Tune the numerical method in $M_{high}$ so that it is reliable at
these draws. All application specific knowledge and classical numerical
analysis can be used here.
3. Compute importance weights $\frac{p_{high}}{p_{low}}$
4. Compute the $\hat{k}$ diagnostic. If $\hat{k} > 0.5$, raise precision of
the numerical method in $M_{low}$ and go back to step 1.
5. Resample draws of $M_{low}$, using importance weights to get draws from
$M_{high}$, and therefore essentially draws from $M$.
:::

The next two sections of this case study outline how to apply this workflow to
do fast but reliable inference for

1. an ODE model using a built-in Stan solver
2. a PDE model using a custom solver that does not have explicit tolerance
controls

The importance sampling diagnostics are handled with the
[`loo`](https://mc-stan.org/loo) package and the resampling is handled with the
[`posterior`](https://github.com/stan-dev/posterior) package.

# SIR example (ODE)

Here we study a classic Susceptible-Infected-Recovered (SIR) model of disease
spread. The code is adapted from that in the Stan Case Study [@grinsztajn2020], 
which provides an introduction to disease
transmission modeling in Stan in general.

For the purposes of this case study, the goal is to use a very low
precision ODE solver to do inference and then check it afterwards against
a high precision solver. This is useful in practice if sampling with the
high precision solver itself would take an inordinate amount of time.

The states of the ODE are amounts of susceptible (S), infected (I) and recovered
(R) people. The dynamics are given by the ODE system:

\begin{align}
\frac{dS}{dt} &= -\beta \cdot I \cdot \frac{S}{N_{pop}} \\
\frac{dI}{dt} &=  \beta  \cdot I \cdot \frac{S}{N_{pop}} - \gamma \cdot I \\
\frac{dR}{dt} &=  \gamma \cdot I,
\end{align}
where $N_{pop}$ is the population size. The parameters $\beta$ and $\gamma$
will be estimated from time series observations of the number of infected
people (I).


```{r sir_model, message=FALSE, results=FALSE}
model <- stan_model("sir.stan")
```

We print the entire Stan code for our model here.
```{r sir_model_code}
cat(model@model_code)
```

## Solving the system

We will import a function from `sir.stan` for solving the ODE.
It uses the `integrate_ode_rk45` function that is built into Stan.
The actual function exposed from the Stan model (`stan_solve_sir`) is a bit
awkward so we rewrap it here in a way that is easier to use.

```{r sir_solve}
expose_stan_functions(model)

# Solve the SIR system
# - theta = c(beta, gamma), parameters
# - opts = c(rtol, atol, max_num_steps), solver options
# - ts = vector of output time points
solve_sir <- function(ts, theta, opts) {
  stan_solve_sir(ts, theta, c(0), opts[1], opts[2], opts[3])
}
```

The RK45 solver in Stan is an adaptive time step solver, which estimates the
local error of the solution and adapts its step size so that the local error
estimates are less than `atol + rtol * abs(y)`, where `y` is the ODE solution,
and `atol` and `rtol` are called absolute and relative tolerance,
respectively. These tolerances need to be given, and affect both the accuracy
and computational cost of the solution. In general, `rtol` is the tolerance on
the relative error the solver can make when `y` is far from zero. When `abs(y)`
is small (of the order of `atol` or smaller), there is no need to achieve the
relative tolerance.

A third control parameter, `max_num_steps`, determines the maximum number of
steps that can be taken to achieve the tolerance. In practice, we have
observed that setting this to a much smaller value than the default can
lower the warmup times of some chains by several orders of magnitude. This can
be because it possibly helps in rejecting or quickly getting out of 
initial low-probability parameter regions, where the ODE solutions blow up and achieving the tolerances would require a much larger number of steps than in
the good parameter region.

We can just pick some options and quickly run and plot a solution
just to get a feel for what the system looks like:

```{r sir_plot}
plot_sir <- function(t, y) {
  y %>%
    as_tibble() %>%
    setNames(c("I")) %>%
    mutate(Day = t) %>%
    ggplot(aes(Day, I)) +
    geom_line() +
    geom_point() +
    ylab("Infected people")
}

ts <- seq(0.1, 16, by = 0.1)
theta_true <- c(1, 0.2) # true parameter values
opts <- c(1e-4, 1e-4, 100)
ys <- solve_sir(ts, theta_true, opts)
plot_sir(ts, ys)
```

## Generating test data

To test fitting our model we will create noisy measurements of the number of infected people (I) at each day. If we're going to generate data from our model we better have an accurate ODE solver, otherwise we're just generating data
from some weird approximate model.

The simplest way to check that an `atol` and `rtol` are suitable is to do
a solve at one tolerance level, repeat the solve at a much smaller (more
precise) tolerance, and then look at the maximum absolute error at any output
point. We will create a function to do this automatically:

```{r sir_check}
check_reliability_sir <- function(theta, opts) {
  ts <- seq(0.1, 16, by = 0.1)
  y_hat <- solve_sir(ts, theta, opts)
  opts_strict <- c(opts[1] / 10, opts[2] / 10, opts[3])
  y_hat_strict <- solve_sir(ts, theta, opts_strict)
  max_abs_err <- max(abs(y_hat - y_hat_strict))
  return(max_abs_err)
}
```

We can study the maximum absolute error compared to a solution with 10 times
smaller tolerances, as a function `tol = atol = rtol`. Value of `max_num_steps` is kept constant, but if a solver should fail to compute a solution in
those steps, an error is thrown and it needs to be increased. 

```{r sir_tols, fig.width=7, fig.height=3.5}
mae_true <- c()
tols <- 10^(-c(1:12))
for (tol in tols) {
  opts <- c(tol, tol, 1e7)
  mae_true <- c(mae_true, check_reliability_sir(theta_true, opts))
}

qplot(tols, mae_true, geom = c("point", "line")) +
  scale_x_log10() +
  scale_y_log10() + ylab("Max. absolute error") + xlab("Tolerance")
```

From this and our prior knowledge of infectious diseases, we assert
that $10^{-6}$ is a good enough value to use for `atol` and `rtol` during
simulation. Certainly we do not expect have a count of the infected
population accurate to $10^{-4}$ people.

We generate the observed number of infected people (cases) at each time point
$t_i$, from a negative binomial distribution with mean equal to the solution of
$S(t_i)$ from the ODE, and dispersion parameter $\phi = 5$.

```{r sir_noise, fig.width=7, fig.height=4.8}
atol <- 1e-6
rtol <- 1e-6
opts <- c(atol, rtol, 1e7)

N <- 16 # number of data points
t_data <- seq(1, N)
dispersion <- 5 # noise parameter for negative binomial
mu <- solve_sir(t_data, theta_true, opts)
y_data <- rnbinom(length(t_data), mu = mu, size = dispersion)

tibble(t = t_data, mu = mu, y = y_data) %>%
  ggplot() +
  geom_line(aes(t, mu), col = "firebrick") +
  geom_point(aes(t, y)) +
  xlab("Day") +
  ylab("Infected people") +
  ggtitle("Simulated data as points \nUnderlying solution as lines")
```

## Likelihood

We also define an R function that computes the likelihood given the data,
parameter values and solver options.
```{r sir_likelihood, fig.width=7, fig.height=4.5, results=FALSE}
# Likelihood function for the SIR model
# - t_data = vector of measurement times
# - y_data = vector of measurements of number of infected people
# - params = parameter vector c(beta, gamma, phi)
# - opts = c(rtol, atol, max_num_steps), solver options
log_likelihood_sir <- function(t_data, y_data, params, opts) {
  theta <- params[1:2]
  phi <- params[3]
  y_hat <- solve_sir(t_data, theta, opts)
  log_lh <- sum(dnbinom(y_data, size = phi, mu = y_hat, log = TRUE))
  return(log_lh)
}
```

## Applying the workflow

As a reminder, our mission in fitting this ODE is to use a low precision
solver. It is always tempting to use low precision solvers when working
with ODEs because they (usually) run faster. The difficulty becomes how to
deal with the coarser approximation. Does the lower precision cause an effect
that matters? If so, can it be corrected and how? These are the questions the
workflow here will allow us to answer.

### Generating draws from $p_{low}$

The first step in the workflow is to take any low precision approximation 
($M_{low}$) and fit the data. Remember, all our numerical methods are
approximations, and so we refer to this model specifically as a low precision
model. We will check it against a higher precision model later. In this case,
we will use `rtol = 1e-4`, `atol = 1e-3`.

```{r sir_fit, fig.width=7, fig.height=4.5, results=FALSE}
opts_low <- c(1e-4, 1e-3, 100)
stan_data <- list(
  N = length(t_data),
  t_data = t_data,
  y_data = y_data,
  rtol = opts_low[1],
  atol = opts_low[2],
  max_num_steps = opts_low[3]
)

fit1 <- rstan::sampling(model,
                       stan_data,
                       seed = rng_seed_stan,
                       cores = 4
)
```


```{r sir_summary}
pars <- c("beta", "gamma", "phi")
draws1 <- get_draws(fit1, pars)
print(fit1, pars = c("beta", "gamma", "phi"))
```

### Developing a reference model

Before we can check if the importance sampling correction is possible, we
need to have a reference model ($M_{high}$) to compare against. That means we
need a version of the model with tolerances such that it is suitably accurate
across all the posterior draws generated from the low precision model.

In this case, `rtol = atol = 1e-6` was accurate enough generating the data,
so let's check if it is accurate in all the draws in this posterior.

```{r sir_tune}
# Check reliability at each draw and add MAEs to data frame
compute_errors_sir <- function(df, opts) {
  num_draws <- nrow(df)
  mae <- c()
  for (i in 1:num_draws) {
    theta <- as.numeric(df[i, c("beta", "gamma")])
    mae_i <- check_reliability_sir(theta, opts)
    mae <- c(mae, mae_i)
  }
  df$mae <- mae
  return(df)
}

opts_high <- c(1e-6, 1e-6, 1e8)
ode_pars <- c("beta", "gamma")
draws1 <- compute_errors_sir(draws1, opts_high)

p1 <- scatter_colored(draws1, "beta", "gamma", "mae")
p2 <- scatter_colored(draws1, "beta", "phi", "mae")
p3 <- scatter_colored(draws1, "gamma", "phi", "mae")
ggarrange(p1, p2, p3)
```

We can plot this as a distribution and see that `rtol = atol = 1e-6` keeps
us under an absolute error of one milliperson. This seems accurate enough.

```{r sir_hist, message=FALSE}
qplot(draws1$mae, geom = "histogram")
```

### Computing importance weights

With the reference model in place, it is time to compute the importance
weights $\frac{p_{high}}{p_{low}}$ for each post-warmup draw. This is simple:
just compute the log density of the reference model and the log
density of the low precision model and take the difference (we work
with the log of the importance ratios $\log p_{high} - \log p_{low}$
for numeric stability).

The hidden downside is that it might take some time to compute the log
densities of the reference model for each draw. It should still be way faster
than sampling with the reference model itself, since we don't need to compute
gradients, HMC trajectories, evaluate proposals that can be rejected, and 
skip the whole warmup period. Therefore we likely won't have to try to do
accurate ODE solves in the ill-behaved parameter regions, where a huge number
of steps would be needed to achieve the tolerances. Another upside is that
the calculations could be done in parallel for each draw.

If the priors are kept the same between the reference and low precision
model, then those can be left out of this calculation (they will cancel).

```{r sir_weights}
# Compute log likelihood ratio for each draw and add to data frame
log_ratios_sir <- function(df) {
  num_draws <- nrow(df)
  log_lh_low <- rep(0, num_draws)
  log_lh_high <- rep(0, num_draws)
  for (i in seq_len(num_draws)) {
    params_i <- as.numeric(df[i, c("beta", "gamma", "phi")])
    log_lh_low[i] <- log_likelihood_sir(t_data, y_data, params_i, opts_low)
    log_lh_high[i] <- log_likelihood_sir(t_data, y_data, params_i, opts_high)
  }
  df$log_ratio <- log_lh_high - log_lh_low
  return(df)
}

draws1 <- log_ratios_sir(draws1)
p1 <- scatter_colored(draws1, "beta", "gamma", "log_ratio")
p2 <- scatter_colored(draws1, "beta", "phi", "log_ratio")
p3 <- scatter_colored(draws1, "gamma", "phi", "log_ratio")
ggarrange(p1, p2, p3)
```

We can plot the log importance ratios and see they are all close to zero
(which means out approximation was not too bad).

```{r sir_weight_hist, message=FALSE}
qplot(draws1$log_ratios, geom = "histogram")
```

### Computing $\hat{k}$ diagnostic

With the importance ratios calculated, we can check if they are usable or not
with the PSIS $\hat{k}$ diagnostic.

```{r sir_pareto}
r_eff1 <- loo::relative_eff(x = exp(-draws1$log_ratio), draws1$chain_id)
psis1 <- loo::psis(draws1$log_ratio, r_eff = r_eff1)
print(psis1$diagnostics)
print(psis1)
```

$\hat{k} < 0.5$, and so importance sampling should be reliable.

### Resampling

At this point we have a weighted set of posterior draws. It
is usually easier to work with a set of draws than a set of weighted draws, so
we can resample our weighted draws to become a set of unweighted draws using
`posterior::resample_draws`. The effective sample size will be slightly
lowered by such a resampling, but unweighted draws are really more convenient
to work with.

Just because it is possible to do an importance sampling correction on a set
of draws does not mean that unweighted statistics on these draws are safe
to use. In this case, the results are not much different, but it should not
be forgotten:

```{r sir_resample}
w1 <- exp(draws1$log_ratio)
draws1_list <- rstan::extract(fit1, c("beta", "gamma", "phi"))
draws_df1 <- posterior::as_draws_df(draws1_list)
resampled_df1 <- posterior::resample_draws(draws_df1, weights = w1)

print(draws_df1 %>% posterior::summarize_draws())
print(resampled_df1 %>% posterior::summarize_draws())
```

# Conclusions

And that is that! Happy approximating!

# Computation environment

```{r computational_environment, echo=FALSE}
sessionInfo()
cat(paste0("Initial rng seed was ", rng_seed_r, "\n"))
cat(paste0("Stan rng seed was ", rng_seed_stan, "\n"))

# Result check
pareto_k_check <- function(res, expected_ok = TRUE) {
  arg_name <- deparse(substitute(res))
  khat <- res$diagnostics$pareto_k
  msg <- paste0("In ", arg_name, ", pareto_k was ", round(khat, 5), "\n")
  cat(msg)
  return(khat)
}
khat1 <- pareto_k_check(psis1)
if (khat1 > 0.5) {
  stop("Results inconsistent with text!")
}
```

# References


