---
title: "Reliable and efficient ODE model inference"
date: "`r Sys.Date()`"
author: "Juho Timonen, Ben Bales, Harri Lähdesmäki and Aki Vehtari"
link-citations: true
output:
  html_document:
    number_sections: true
    toc: true
    toc_float: true
    toc_depth: 4
    highlight: pygments
    theme: cosmo
    css: "style.css"
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load, message=FALSE}
# Requirements
library(cmdstanr)
library(posterior)
library(bayesplot)
library(ggplot2)
library(loo)
library(stats)
library(outbreaks)
color_scheme_set("brightblue")

rng_seed_r <- 999 # rng seed for MCMC
rng_seed_stan <- 123 # rng seed for data simulation
set.seed(rng_seed_r)
SIG_FIGS <- 10
print_lines <- function(file) {cat(readLines(file), sep="\n")}
```

# Introduction

## ODEs

Ordinary differential equations (ODEs) 

\begin{equation}
\frac{\text{d} \textbf{x}(t)}{\text{d} t} = f_{\theta}(\textbf{x}(t), t)
\end{equation}

are a model for the rate of change in some quantities $\textbf{x}(t)$, and are used in many fields of science. It is common that we have time-course data $\{(t_n, \textbf{y}_n)\}_{n=1}^N$, observed at time points $t_n \neq t_0$, and 
we need to calculate $\textbf{x}(t_n)$ for each $n=1, \ldots, N$, in order to
compute likelihood of the data. Given initial value  $\textbf{x}(t_0) = \textbf{x}_0$, and assuming some smoothness requirements for $f_{\theta}$, there exists a unique solution for $\textbf{x}(t)$, $t \in \mathbb{R}$. This solution is given by the integral formula

\begin{equation}
\textbf{x}(t) = \textbf{x}_0 + \int_{t_0}^t f_{\theta}(\textbf{x}(\tau), \tau) 
\text{d} \tau,
\end{equation}

but the integral usually doesn't have a closed form, and numerical methods are required.

## Example: SIR

Here we study a classic Susceptible-Infected-Recovered (SIR) model of disease
spread. The code is adapted from that in the Stan Case Study [@grinsztajn2020], 
which provides an introduction to disease
transmission modeling in Stan in general. The states of the ODE are amounts of susceptible (S), infected (I) and recovered
(R) people. The dynamics are given by the ODE system:

\begin{align}
\frac{dS}{dt} &= -\beta \cdot I \cdot \frac{S}{N_{pop}} \\
\frac{dI}{dt} &=  \beta  \cdot I \cdot \frac{S}{N_{pop}} - \gamma \cdot I \\
\frac{dR}{dt} &=  \gamma \cdot I,
\end{align}
where $N_{pop}$ is the population size. Using the above notation, we have
$\textbf{x}(t) = [S(t), I(t), R(t)]^{\top}$ and $\theta = \{\beta, \gamma\}$. 

### Stan code

This system can be coded as the following Stan function.

```{r sir_stan_function}
print_lines("sir/functions.stan")
```

Other model parts that we will use are

```{r sir_stan_blocks}
print_lines("sir/data.stan")
print_lines("sir/tdata.stan")
print_lines("sir/params.stan")
print_lines("sir/tparams.stan")
print_lines("sir/prior.stan")
```

Prior
```{r sir_prior_model}
mod_prior <- cmdstan_model("sir/sir_prior.stan", include_paths = "sir")
mod_prior$print()
```

Simulator model
```{r sir_gen_model}
mod_sim <- cmdstan_model("sir/sir_simulate.stan", include_paths = "sir")
mod_sim$print()
```

Posterior
```{r sir_post_model}
mod_post <- cmdstan_model("sir/sir_posterior.stan", include_paths = "sir")
mod_post$print()
```

### Data

```{r sir_data}
y <- influenza_england_1978_school$in_bed
N <- length(y)
data_list <- list(
  y = y,
  N = N,
  t = seq(1, N),
  I0 = 1,
  pop_size = 763
)
print(data_list)
```


### Prior sampling
```{r sample_prior}
# Sample the prior and extract draws
prior_fit <- mod_prior$sample(
  sig_figs = SIG_FIGS, 
  seed = rng_seed_stan,
  refresh = 1000
)
print(prior_fit)
prior_draws <- prior_fit$draws(c("beta", "gamma", "phi_inv"))
mcmc_areas(prior_draws)
```


### Prior checking


Prior check
```{r prior_check}
# Function for simulating ODE solutions and data given parameter( draws)s
simulate_sir <- function(params, t_sim=NULL, RTOL = 1e-6, ATOL = 1e-6,
                         MAX_NUM_STEPS = 1e6) {
  if(is.null(t_sim)) t_sim <- data_list$t
  stan_data <- list(
    t = t_sim,
    N = length(t_sim),
    RTOL = RTOL,
    ATOL = RTOL,
    MAX_NUM_STEPS = MAX_NUM_STEPS,
    pop_size = data_list$pop_size,
    I0 = data_list$I0
  )
  sim <- mod_sim$generate_quantities(
    data = stan_data, 
    fitted_params = params
  )
}

prior_sim <- simulate_sir(prior_draws)
x_prior_sim <- prior_sim$draws("x")

# Plot
plot(data_list$t, rep(data_list$pop_size, data_list$N), type="l", lty=2,
     col="gray70", ylim=c(0,800), ylab="Infected", xlab="Day")
for(i_chain in 1:4){
  for(i_draw in 1:100) {
    S <- as.vector(x_prior_sim[i_draw, i_chain, 1:N])
    I <- as.vector(x_prior_sim[i_draw, i_chain, (N+1):(2*N)])
    R <- as.vector(x_prior_sim[i_draw, i_chain, (2*N+1):(3*N)])
    lines(data_list$t, I, col="firebrick")
  }
}
points(data_list$t, y, ylim = c(0, 1000), pch=20)
```


## Computational challenges

When performing Bayesian inference for the parameters $\theta$
(and possible other model parameters, note that also $\textbf{y}_0$ can be a
parameter) using MCMC methods, the system needs to be solved numerically on each log posterior probability evaluation. Furthermore, when using a gradient
based MCMCM algorithm like Stan does, also _sensitivities_, i.e. 
derivatives of the solution w.r.t. parameters, are needed. This is done
in Stan by numerically solving an extended ODE system consisting of the
sensitivities and the original dynamical variables. 

The solution provided by a numerical method
is always an approximation to the true solution. 
This means that the computed posterior probability density and its
gradient are also approximations and the whole MCMC inference is biased
to some degree. So we are performing Bayesian inference for the parameters, the error in the solutions propagates also to the statistical properties of the posterior draws, in ways that cannot be predicted beforehand.

How could this error be studied? If the model parameters are fixed,
we can verify that the solution at all points is
appropriately close to a more accurate reference solution. 
There is a vast amount of classical literature that studies this error, and the ODE solvers built into
Stan are these classical methods that obtain estimate their error and adapt
their step size so that a given tolerance is achieved. However, we are doing statistics, and so we need to
know that the solution is accurate enough across all relevant parts of
parameter space. Additionally, it is not known beforehand where the 
"relevant parts of parameter space" are!

The problem of validating the use of a numerical method for a Bayesian model
is therefore significantly more complicated than in the classical numerical
analysis world. One idea is to run sampling repeatedly, gradually increasing
the solver accuracy, until the posterior inference
results don't change anymore. But this might not
be computationally very attractive. The numerical methods are often expensive
and frequently become the limiting factors in whatever model they are involved in. Moreover, achieving more accuracy often comes with the cost of more
computation time. Note however that there is no direct tradeoff.

These two problems (computational cost and potential unreliability) occur also
with any other type of model that requires numerically solving an implicitly defined function or variable, but in this case study we focus on ODEs.

The point of this case study is to show how by adding one
additional tool, namely Pareto-Smoothed Importance Sampling (PSIS) [@yao2018;
@vehtari2019], we can solve this problem.


## Simpler numerical methods
Simplest numerical methods for integrating ODEs usually involve some
discretization in time, which affects the solver accuracy. Denser 
discretization means more accuracy but also more computation, as more time
steps need to be taken.

The simplest things we might do to speed up our calculations are lowering
the timestep, coarsening the discretization, or increasing the tolerance of the
solvers. That immediately leaves us with the question, is this okay? Has
changing the numerical method affected our parameter inference results? Was our
original method giving correct inference results to begin with? Are the default
tolerances in Stan suitable for the problem at hand?

# Workflow

Let $M$ be the model for which we would like to perform inference, but which we
cannot evaluate since the likelihood is defined implicitly through an
ODE or PDE system that is not analytically tractable. MCMC inference for $M$
can be seen actually as inference for another model $M_{high}$, which
is the same model as $M$ but using a numerical solver, and can therefore be
evaluated.

Our workflow addresses the problem of defining the high-precision numerical
method in $M_{high}$ so that $M_{high}$ can trusted to have essentially the
same posterior as $M$. We define a way to perform inference for $M_{high}$ 
without needing to compute gradients or HMC trajectories for it. This involves
another model $M_{low}$, which is again the same model, except that $M_{low}$
uses a cheaper and less accurate numerical methods (or just looser tolerances
and/or coarser discretization) to compute the required ODE or PDE solutions,
and is therefore faster to fit. The posterior densities are denoted $p_{low}$
and $p_{high}$, respectively.

To understand how PSIS comes into play, we must first discuss importance
sampling. If we want to compute expectations with the high precision model, we
can take draws from the low precision models and reweight these according to the
importance weights $\frac{p_{high}}{p_{low}}$. If these models are too
different, then the reweighting will produce noisy estimates that are not
useful. PSIS and particularly the Pareto $k$-diagnostic (denoted $\hat{k}$), is
the tool that tells us when we can or cannot rely on the importance weights. If
$\hat{k} < 0.5$ we are safe to do the importance sampling, if $\hat{k} < 0.7$
the importance sampling will start to converge more slowly, and if
$\hat{k} > 0.7$ the importance sampling estimates are unreliable. For simplicity
we will only consider the $\hat{k} < 0.5$ threshold.

Ideally, $M_{high}$ would involve a numerical method that we can trust
completely in all parts of the parameter space so that, as long as
$\hat{k} < 0.5$, importance weights can be used to reweight the low precision
approximation $p_{low}$ to the high precision approximation $p_{high}$. We can
think of $M_{high}$ as a reference model, because it is the baseline to which
we compare. It is difficult in practice to know if a given model is a good
reference model in all parts of parameter space, due to curse of dimensionality
and the fact that analysed system can have different properties in different
parts of the parameter space. For example, ODEs can qualitatively change their
behaviour as a function of parameters (bifurcation), or become stiff or
chaotic in some parameter regions. Accuracy can be checked at a given set of
parameters fairly easily, but not over a high dimensional parameter space.
Under these conditions it is necessary to compromise to develop a reference
model that works only over a range of parameter space, but even then it is hard
to know *a priori* what range that is.

We propose the following workflow:

::: {#workflow .box}
1. Generate draws from $p_{low}$.
2. Tune the numerical method in $M_{high}$ so that it is reliable at
these draws. All application specific knowledge and classical numerical
analysis can be used here.
3. Compute importance weights $\frac{p_{high}}{p_{low}}$
4. Compute the $\hat{k}$ diagnostic. If $\hat{k} > 0.5$, raise precision of
the numerical method in $M_{low}$ and go back to step 1.
5. Resample draws of $M_{low}$, using importance weights to get draws from
$M_{high}$, and therefore essentially draws from $M$.
:::

The next two sections of this case study outline how to apply this workflow to
do fast but reliable inference for

1. an ODE model using a built-in Stan solver
2. a PDE model using a custom solver that does not have explicit tolerance
controls

The importance sampling diagnostics are handled with the
[`loo`](https://mc-stan.org/loo) package and the resampling is handled with the
[`posterior`](https://github.com/stan-dev/posterior) package.


## Example: SIR fitting
The parameters $\beta$ and $\gamma$
will be estimated from time series observations of the number of infected
people (I). For the purposes of this case study, the goal is to use a very low
precision ODE solver to do inference and then check it afterwards against
a high precision solver. This is useful in practice if sampling with the
high precision solver itself would take an inordinate amount of time.


## References